The function--sheaf dictionary for smooth commutative group schemes over finite fields, with applications over local fields

### 2014-08-21 dictionary.tex
Small changes

\date{\today}
\author{Clifton Cunningham}
\email{cunning@math.ucalgary.ca}
\address{University of Calgary}
\author{David Roe}
\email{roed.math@gmail.com}
\address{Pacific Institute for the Mathematical Sciences, University of Calgary}
\subjclass[2010]{14F05 (primary), 14L15 (secondary)}
\keywords{function--sheaf dictionary, trace of Frobenius, geometrization, quasicharacters, quasicharacter sheaves, smooth commutative group schemes}

@@

In this paper we find the group $\QCiso{G}$ [point]
and also the set $\Hom(\qcs{L},\qcs{L}')$, for all quasicharacter sheaves $\qcs{L}$, $\qcs{L}'$ on $G$ [point]; in this way we provide a complete description of the category $\QC(G)$.

\begin{remark}
The category of quasicharacter sheaves on $G$ is not abelian; $\QC(G)$
is not a tensor category in the sense of \cite{deligne:02a}*{0.1}.
\end{remark}

@@

Let $K$ be a local field (in the sense of Serre) with residue field $\Fq$. 
In this section we consider connected smooth commutative algebraic groups over $K$ that admit a Néron model, by which we means a perhaps lft Néron model.
By \cite{bosch-lutkebohmert-reynaud:NeronModels}*{\S 10.2, Thm 2}, these are precisely the connected smooth commutative algebraic groups over $K$ that contain no subgroup of type $\mathbb{G}_\text{a}$.
We remark that these connected commutative algebraic groups over $K$ form an additive category in a natural way, as a full subcategory of connected commutative algebraic groups over $K$, and that this category includes algebraic tori over $K$ and Abelian varieties over $K$.

Let $R$ be the ring of integers of $K$.
Write $\m$ for the maximal ideal of $R$ and set $R_n = R/\m^{n+1}$, for every non-negative integer $n$.
Let $X_K$ be a connected smooth commutative algebraic groups over $K$ that admit a Néron model, $X$.
A {\it quasicharacter of $X_K(K)$} is a group homomorphism $X_K(K) \to \EEx$ that factors through $X_K(X) \to X(R_n)$, for some non-negative integer $n$;
the group of quasicharacters of $X_K(K)$ will be denoted by $\Hom_\text{qc}(X_K(K),\EEx)$ and the subgroup of those that factor through $X(R_n)$ will be denoted by $\Hom_n(X_K(K),\EEx)$.
%
In this section we explain how to geometrize and categorify quasicharacters of $X_K(K)$ using quasicharacter sheaves.

@@

\begin{theorem}\label{thm:QCXK}
Let $K$ and $R$ be as above.
The trace of Frobenius provides a natural transformation between the additive functors
\[
F_{\QC}: X_K \mapsto \QCiso{\Gr_R(X)}
\qquad\text{and}\qquad
F_{\text{qc}}: X_K \mapsto \Hom_{\text{qc}}(X_K(K),\EEx)
\]
where $X$ is a Neron model for $X_K$, as functors from the additive category of connected commutative algebraic groups over $K$ that admit Neron models to the category of abelian groups. 
Regarding this natural transformation:
\begin{enumerate}
\item[(1)] there is a canonical short exact sequence of abelian groups 
\[
0 \to \Hh^2(\pi_0(X)_{\bFq},\EEx)^{\Frob{}} \to \QCiso{\Gr_R(X)} \to \Hom_\text{qc}(X_K(K),\EEx) \to 0;
\] 
\item[(2)] for all quasicharacter sheaves $\qcs{F}$, $\qcs{F}'$ on $\Gr_R(X)$, and for every $\rho \in \Hom(\qcs{F},\qcs{F}')$, either $\rho$ is trivial or $\rho$ is an isomorphism;
\item[(3)] for all quasicharacter sheaves $\qcs{F}$ on $\Gr_R(X)$,
\[
\Aut(\qcs{F}) \iso \Hom((\pi_0(X)_{\Fq})_{\Weil{}},\EEx),
\]
canonically.
\end{enumerate}
\end{theorem}

@@

\subsection{Natural transformation}

>>>>

\date{\today}
\author{Clifton Cunningham}
\email{cunning@math.ucalgary.ca}
\address{University of Calgary}
\author{David Roe}
\email{roed.math@gmail.com}
\address{Pacific Institute for the Mathematical Sciences, University of Calgary}
\subjclass[2010]{14F05 (primary), 14L15 (secondary)}
\keywords{function--sheaf dictionary, geometrization, quasicharacters, quasicharacter sheaves, smooth commutative group schemes}

@@

We will describe the group $\QCiso{G}$ in Theorem \ref{thm:snake}
and the sets $\Hom(\qcs{L},\qcs{L}')$ in Theorem \ref{thm:autornaught}; in this way we provide a complete description of the category $\QC(G)$.

\begin{remark}
The category of quasicharacter sheaves on $G$ is not abelian since it is not closed under direct sums and thus $\QC(G)$
is not a tensor category in the sense of \cite{deligne:02a}*{0.1}.  We suspect that requiring that $\mu$ be injective rather than
an isomorphism and dropping the condition that the stalks be one-dimensional would yield an abelian category.
\end{remark}

@@

In this section we use quasicharacter sheaves to geometrize and categorify the group 
of admissible characters of smooth commutative group schemes over local fields that admit Néron models.

Let $R$ be a complete discrete valuation ring $R$ with residue field $\Fq$.
Write $\m$ for the maximal ideal of $R$ and $K$ for the quotient field for $R$.
For every non-negative integer $n$,  set $R_n = R/\m^{n+1}$.

Suppose $X_K$ is a connected commutative algebraic group over $K$ that admits a Néron model $X$ over $R$.
%
A {\it quasicharacter of $X_K(K)$} is a group homomorphism $X_K(K) \to \EEx$ that factors through $X_K(X) \to X(R_n)$, for some $n\in\NN_0$;
the group of quasicharacters of $X_K(K)$ will be denoted by $\Hom_\text{qc}(X_K(K),\EEx)$ and the subgroup of those that factor through $X(R_n)$ will be denoted by $\Hom_n(X_K(K),\EEx)$.
%
In this section we explain how to geometrize and categorify quasicharacters of $X_K(K)$ using quasicharacter sheaves.

It seems likely that there connected commutative algebraic groups over local fields that admit Neron models but are neither algebraic tori nor Abelian varieties; indeed, it is conjectured that K-wound unipotent group schemes over local fields like $\Fq((t))$ are examples. We have use the ponderous phrase "let $X_K$ be a connected commutative algebraic group over $K$ that admits a Néron model $X$" in Section~\ref{sec:applications} to draw attention to the fact that our theory of quasicharacter sheaves is applicable to more than algebraic tori and Abelian varieties.

@@

\begin{theorem}\label{thm:QCXK}
Let $X_K$ be a commutative algebraic group over $K$ with Néron model $X$ over $R$. 
\begin{enumerate}
\item[(1)] There is a canonical short exact sequence of abelian groups 
\[
0 \to \Hh^2(\pi_0(X)_{\bFq},\EEx)^{\Frob{}} \to \QCiso{\Gr_R(X)} \to \Hom_\text{qc}(X_K(K),\EEx) \to 0.
\] 
\item[(2)] For all quasicharacter sheaves $\qcs{F}$, $\qcs{F}'$ on $\Gr_R(X)$, and for every $\rho \in \Hom(\qcs{F},\qcs{F}')$, either $\rho$ is trivial or $\rho$ is an isomorphism.
\item[(3)] For all quasicharacter sheaves $\qcs{F}$ on $\Gr_R(X)$,
\[
\Aut(\qcs{F}) \iso \Hom((\pi_0(X)_{\Fq})_{\Weil{}},\EEx),
\]
canonically.
\end{enumerate}
\end{theorem}

@@

\subsection{Group schemes that admit a Neron model}
